Carbonate chemistry at Ny-Ålesund

### 2021-03-12 awipev-CO2.Rmd
corrected depth in stratification; added many rm for large objects to clear memory

```{r read observatory data, cleanup and prepare ferrybox data, echo=FALSE, warning=FALSE, message=FALSE}
d_hour <-
  readRDS(file = paste0(
    path,
    "fb_awipev-co2_server/ny-alesund/data/NRT_data/nydata_hour.rds"
  ))
d_hour <- d_hour %>%
  dplyr::filter(datetime <= as.POSIXct("2021-01-10")) %>% #cutoff date
  dplyr::select(-pco2_inst) %>% # this is the serial number of pCO2 sensor, not needed
  dplyr::rename(
    pco2_fb = pco2_corr_filtered,
    temp_dur = temp_dur_filtered,
    pco2_raw = pco2_raw_filtered,
    sal_fb = sal_fb_filtered,
    temp_insitu_11m = temp_insitu_11m_filtered,
    temp_fb = temp_fb_filtered,
    ph_dur = ph_dur_filtered
  ) %>% 
  arrange(datetime)
d_all <- as_tibble(d_hour) #d_all will be the merger
d_all <- d_all %>%
  dplyr::mutate_all(function(x) ifelse(is.nan(x), NA, x)) %>% # replace all NaN by NA
  dplyr::mutate(depth = ifelse( # create new variable depth
      pressure_insitu_ctd <= 2,
      1,
      ifelse(
        pressure_insitu_ctd > 2 & pressure_insitu_ctd <= 4,
        3,
        ifelse(
          pressure_insitu_ctd > 4 & pressure_insitu_ctd <= 6,
          5,
          ifelse(
            pressure_insitu_ctd > 6 & pressure_insitu_ctd <= 8,
            7,
            ifelse(pressure_insitu_ctd > 8, 9, NA)
          )
        )
      )
    )) %>%
  dplyr::rowwise() %>% 
  dplyr::mutate(n_depth_ctd = sum(is.na( # create new variable to identify the hours for which there are CTD data at more than 1 depth 
  c(
    sal_insitu_ctd_1m,
    sal_insitu_ctd_3m,
    sal_insitu_ctd_5m,
    sal_insitu_ctd_7m,
    sal_insitu_ctd_9m
  )
), na.rm = TRUE)) %>% 
  ungroup()
#restore datetime, date which are mangled by the mutates above, arrange ascending
d_all <- d_all %>% 
  dplyr::mutate(datetime = d_hour$datetime, 
                date = d_hour$date,
                year = lubridate::year(d_hour$datetime), # year as a number
                monthd = lubridate::month(d_hour$datetime), # month as a number
                montht = lubridate::month(d_hour$datetime, label = TRUE, abbr = TRUE) # abbreviated months
  )
```

@@

# we look at 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, pHint_tot_sf, depth) %>%
  dplyr::mutate(depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 9) %>% 
  group_by(depth, montht)
summary(aov(tmp$pHint_tot_sf ~ tmp$depth * tmp$montht))

@@

@@

```{r stratification, echo=FALSE, message= FALSE, warning=FALSE}
# we look at salinity all depths
tmp <- d_all %>% 
  dplyr::select(montht, sal_insitu_ctd, depth) %>%
  group_by(depth, montht)
p <- tmp %>% 
  ggplot(aes(x=as.factor(depth), y = sal_insitu_ctd)) +
  geom_boxplot(notch=FALSE, size=0.2, fill="grey50", outlier.color = "blue", outlier.size = 0.3) +
  labs(title = "Salinity vs depth", x = "Depth (m)", y= "Salinity") +
  facet_wrap(vars(montht), scale = "free_y")
print(p)

# heatmap
tmp <- d_all %>%
  dplyr::filter(!is.na(depth)) %>%
  dplyr::select(montht, sal_insitu_ctd, depth) %>%
  group_by(montht, depth)
sal_median <- skim(tmp) %>%
  dplyr::select(depth, montht, numeric.p50) %>%
  pivot_wider(names_from = depth, values_from = c(numeric.p50))

plt <- sal_median %>%
  ggplot(aes(
    x = montht,
    y = as.factor(depth),
    fill = median(sal_insitu_ctd)
  ), color = "white") +
  geom_tile() +
  scale_fill_gradientn(colors = viridis(5), breaks = seq(31, 36, by = 1), na.value = "grey") +
  labs(x = "", y = "", fill = "Salinity") +
  scale_x_discrete(position = "top") +
  scale_y_discrete(name = "", limits = rev(levels(as.factor(tmp$depth)))) +
  # scale_y_discrete(name = "", limits = rev(levels(as.factor(tmp$depth))),
  #                  rev(levels(tmp$depth))[c(1,3,5,7,9)]) +
  # coord_equal() +
  theme(legend.position = "top",
        axis.text.y = element_text(size = 13),
        axis.text.x = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.title = element_text(size = 14),
        #legend.key.size = unit(1.5, "cm"),
        legend.key.height = unit(0.8, "cm"),
        legend.key.width  = unit(2, "cm"))
#plt

# we look at salinity 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, sal_insitu_ctd, depth) %>%
  dplyr::mutate(depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 9) %>% 
  group_by(depth, montht)
sal_range <- skim(tmp) %>% 
  dplyr::select(depth, montht, numeric.p50) %>% 
  pivot_wider(names_from = depth, values_from = c(numeric.p50)) %>% 
  dplyr::rename(median2m = "2",
                median9m = "9") %>% 
  dplyr::mutate(sal_range = median2m - median9m)
kable(sal_range, digits = 2, caption = "Median salinity")

# we look at temperature 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, temp_insitu_ctd, depth) %>%
  dplyr::mutate(depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 9) %>% 
  group_by(depth, montht)
temp_range <- skim(tmp) %>% 
  dplyr::select(depth, montht, numeric.p50) %>% 
  pivot_wider(names_from = depth, values_from = c(numeric.p50)) %>% 
  dplyr::rename(median2m = "2",
                median9m = "9") %>% 
  dplyr::mutate(temp_range = median2m - median9m)
kable(temp_range, digits = 2, caption = "Median temperature")
summary(aov(tmp$temp_insitu_ctd ~ tmp$depth * tmp$montht))

p <- tmp %>% 
  ggplot(aes(x=as.factor(depth), y = temp_insitu_ctd)) +
  geom_boxplot(notch=FALSE, size=0.2, fill="grey50", outlier.color = "blue", outlier.size = 0.3) +
  labs(title = "Temperature vs depth", x = "Depth (m)", y= "Salinity") +
  facet_wrap(vars(montht), scale = "free_y")
print(p)

# we look at density 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, sal_insitu_ctd, temp_insitu_ctd, depth) %>%
  dplyr::mutate(density = seacarb::rho(S = sal_insitu_ctd, T = temp_insitu_ctd, P = 0),
                depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 9) %>% 
  dplyr::select(montht, depth, density) %>% 
  group_by(depth, montht) 
den_range <- skim(tmp) %>% 
  dplyr::select(depth, montht, numeric.p50) %>% 
  pivot_wider(names_from = depth, values_from = c(numeric.p50)) %>% 
  dplyr::rename(median2m = "2",
                median9m = "9") %>% 
  dplyr::mutate(den_range = median2m - median9m)
kable(den_range, digits = 2, caption = "Median density")
summary(aov(tmp$density ~ tmp$depth * tmp$montht))

p <- tmp %>% 
  ggplot(aes(x=as.factor(depth), y = density)) +
  geom_boxplot(notch=FALSE, size=0.2, fill="grey50", outlier.color = "blue", outlier.size = 0.3) +
  labs(title = "Density vs depth", x = "Depth (m)", y= "Salinity") +
  facet_wrap(vars(montht), scale = "free_y")
print(p)
```
Using the medians because data are missing.

>>>>

```{r read observatory data, cleanup and prepare ferrybox data, echo=FALSE, warning=FALSE, message=FALSE}
d_hour <-
  readRDS(file = paste0(
    path,
    "fb_awipev-co2_server/ny-alesund/data/NRT_data/nydata_hour.rds"
  ))
d_hour <- d_hour %>%
  dplyr::filter(datetime <= as.POSIXct("2021-01-10")) %>% #cutoff date
  dplyr::select(-pco2_inst) %>% # this is the serial number of pCO2 sensor, not needed
  dplyr::rename(
    pco2_fb = pco2_corr_filtered,
    temp_dur = temp_dur_filtered,
    pco2_raw = pco2_raw_filtered,
    sal_fb = sal_fb_filtered,
    temp_insitu_11m = temp_insitu_11m_filtered,
    temp_fb = temp_fb_filtered,
    ph_dur = ph_dur_filtered
  ) %>% 
  arrange(datetime)
d_all <- as_tibble(d_hour) #d_all will be the merger
d_all <- d_all %>%
  dplyr::mutate_all(function(x) ifelse(is.nan(x), NA, x)) %>% # replace all NaN by NA
  dplyr::mutate(depth = ifelse( # create new variable depth
      pressure_insitu_ctd <= 2,
      1,
      ifelse(
        pressure_insitu_ctd > 2 & pressure_insitu_ctd <= 4,
        3,
        ifelse(
          pressure_insitu_ctd > 4 & pressure_insitu_ctd <= 6,
          5,
          ifelse(
            pressure_insitu_ctd > 6 & pressure_insitu_ctd <= 8,
            7,
            ifelse(pressure_insitu_ctd > 8, 10, NA)
          )
        )
      )
    )) %>%
  dplyr::rowwise() %>% 
  dplyr::mutate(n_depth_ctd = sum(is.na( # create new variable to identify the hours for which there are CTD data at more than 1 depth 
  c(
    sal_insitu_ctd_1m,
    sal_insitu_ctd_3m,
    sal_insitu_ctd_5m,
    sal_insitu_ctd_7m,
    sal_insitu_ctd_9m
  )
), na.rm = TRUE)) %>% 
  ungroup()
#restore datetime, date which are mangled by the mutates above, arrange ascending
d_all <- d_all %>% 
  dplyr::mutate(datetime = d_hour$datetime, 
                date = d_hour$date,
                year = lubridate::year(d_hour$datetime), # year as a number
                monthd = lubridate::month(d_hour$datetime), # month as a number
                montht = lubridate::month(d_hour$datetime, label = TRUE, abbr = TRUE) # abbreviated months
  )
```

@@

# we look at 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, pHint_tot_sf, depth) %>%
  dplyr::mutate(depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 10) %>% 
  group_by(depth, montht)
summary(aov(tmp$pHint_tot_sf ~ tmp$depth * tmp$montht))

@@

## Stratification
```{r stratification, echo=FALSE, message= FALSE, warning=FALSE}
depth_labels <- c("0-2", "2-4", "4-6", "6-8", "8-12")
# we look at salinity all depths
tmp <- d_all %>% 
  dplyr::select(montht, sal_insitu_ctd, depth) %>%
  group_by(depth, montht)
p <- tmp %>% 
  ggplot(aes(x=as.factor(depth), y = sal_insitu_ctd)) +
  geom_boxplot(notch=FALSE, size=0.2, fill="grey50", outlier.color = "blue", outlier.size = 0.3) +
  labs(title = "Salinity vs depth", x = "Depth (m)", y= "Salinity") +
  scale_x_discrete(labels = depth_labels) +
  facet_wrap(vars(montht), scale = "free_y")
print(p)

# we look at salinity 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, sal_insitu_ctd, depth) %>%
  dplyr::mutate(depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 10) %>% 
  group_by(depth, montht)
sal_range <- skim(tmp) %>% 
  dplyr::select(depth, montht, numeric.p50) %>% 
  pivot_wider(names_from = depth, values_from = c(numeric.p50)) %>% 
  dplyr::rename(median2m = "2",
                median10m = "10") %>% 
  dplyr::mutate(sal_range = median2m - median10m)
kable(sal_range, digits = 2, caption = "Median salinity")

# we look at temperature 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, temp_insitu_ctd, depth) %>%
  dplyr::mutate(depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 10) %>% 
  group_by(depth, montht)
temp_range <- skim(tmp) %>% 
  dplyr::select(depth, montht, numeric.p50) %>% 
  pivot_wider(names_from = depth, values_from = c(numeric.p50)) %>% 
  dplyr::rename(median2m = "2",
                median10m = "10") %>% 
  dplyr::mutate(temp_range = median2m - median10m)
kable(temp_range, digits = 2, caption = "Median temperature")
summary(aov(tmp$temp_insitu_ctd ~ tmp$depth * tmp$montht))

p <- tmp %>% 
  ggplot(aes(x=as.factor(depth), y = temp_insitu_ctd)) +
  geom_boxplot(notch=FALSE, size=0.2, fill="grey50", outlier.color = "blue", outlier.size = 0.3) +
  labs(title = "Temperature vs depth", x = "Depth (m)", y= "Temperature") +
  facet_wrap(vars(montht), scale = "free_y")
print(p)

# we look at density 0 to 4 m vs more than 8 m
tmp <- d_all %>% 
  dplyr::select(montht, sal_insitu_ctd, temp_insitu_ctd, depth) %>%
  dplyr::mutate(density = seacarb::rho(S = sal_insitu_ctd, T = temp_insitu_ctd, P = 0),
                depth = as.factor(ifelse(depth == 1 | depth == 3, 2, depth))) %>% 
  dplyr::filter(depth == 2 | depth == 10) %>% 
  dplyr::select(montht, depth, density) %>% 
  group_by(depth, montht) 
den_range <- skim(tmp) %>% 
  dplyr::select(depth, montht, numeric.p50) %>% 
  pivot_wider(names_from = depth, values_from = c(numeric.p50)) %>% 
  dplyr::rename(median2m = "2",
                median10m = "10") %>% 
  dplyr::mutate(den_range = median2m - median10m)
kable(den_range, digits = 2, caption = "Median density")
summary(aov(tmp$density ~ tmp$depth * tmp$montht))

p <- tmp %>% 
  ggplot(aes(x=as.factor(depth), y = density)) +
  geom_boxplot(notch=FALSE, size=0.2, fill="grey50", outlier.color = "blue", outlier.size = 0.3) +
  labs(title = "Density vs depth", x = "Depth (m)", y= "Density") +
  scale_x_discrete(labels = c("0-4", "8-12")) +
  facet_wrap(vars(montht), scale = "free_y")
print(p)
```
Using the medians because data are missing.

@@
